---
title: "How to Write Custom Functions in R"
author: jpiaskowski
date: 2021-11-18
categories: ["R"]
tags: ["function", "source code"]
draft: true
---

 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE, include=FALSE)
```


Super simple functions
```{r}
my_func1 <- function(x) {
  x^2
}

my_func2 <- function(y) {
  y + 1
}
```
Call the function:
```{r}
x1 <- 10
x2 <- 1:5
x3 <- rnorm(5)

my_func(x1)
my_func(x2)
my_func(x3)
```
Note that it works with data of all sizes. 

Assign the function:
```{r}
(rsquares <- my_func2(x3))
```

Build a more complex function
```{r}
my_boxplot_func <- function(df, yvar, xvar, col){
  boxplot(yvar ~ xvar, data = df, col = col)
}
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

