---
title: "How to Find Function Source Code in R"
authors:
- jpiaskowski
categories: ["R"]
draft: true
date: "2021-11-17T00:00:00Z"
image:
  caption: "Codex Sinaiticus Syriacus"
  focal_point: ""
lastMod: "2021-11-17T00:00:00Z"
projects: []
subtitle: 
summary: 
tags: ["source code", "pryr"]
---

---
title: "How to Find Function Source Code in R"
author: jpiaskowski
date: 2021-11-18
categories: ["R"]
tags: ["function", "source code"]
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Source code in the console

Sometimes this is an easy as typing the function name into the console:

```{r}
lm
```

But, this fails often

```{r}
mean
```

or, it works, but the functon calls another internal function that you can't access easily

```{r}
var
```



what methods are out there?
```{r}
methods(mean)
```

```{r}
mean.default
```
by looking at this, you can see there is considerable error checking. If you are confident that error checking is not needed for a particular data set, you can rewrite the function to skip those steps. 

This one is really handy
```{r}
methods(class = "aov")
```

Much of R source code is written in C (.Internal and .Primitive calls). As such, it's likely to be very efficient already. Altering this is difficult. But, you can  look for that code with this function:

This will search github and open a browser where that code might be
```{r}
library(pryr)
show_c_source(.Internal(mean(x)))
```

honestly, one of best ways to find R code is 






